---
title: "Expresiones regulares"
#subtitle: "⚔<br/>with xaringan"
#author: "Yihui Xie"
#institute: "RStudio, Inc."
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

background-image: url(https://upload.wikimedia.org/wikipedia/commons/b/be/Sharingan_triple.svg)

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(dplyr)
library(ggplot2)
library(forcats)
library(stringr)
```

???

Image credit: [Wikimedia Commons](https://commons.wikimedia.org/wiki/File:Sharingan_triple.svg)

---
class: center, middle

# xaringan

### /ʃaː.'riŋ.ɡan/

---
class: inverse, center, middle

# Get Started

---

# RegEx

Las expresiones regulares son representaciones esquemáticas o semi esquemáticas de patrones.

--

- CV.CV.CV

--

- FN(FV-FN)

--

- Ca.Cá

--

- FN(DAR-FN-FP)

---
background-image: url(`r xaringan:::karl`)
background-position: 50% 50%
class: center, bottom, inverse

# You only live once!

---

# RegEx

Las expresiones regulares se utilizan principalmente para la identificación de patrones en cadenas de carácteres.

```{r}

string <- c("Handel", "Händel", "Haendel", "banana", "anana", "piña", "piñata")

str_subset(string, "H(a|ä|(ae))ndel")

str_subset(string, "\\w+nana")

```

---

# RegEx en un corpus

En un corpus, el uso de RegEx es útil para identificar contextos relevantes para el investigador.

```{r}

archivo <- "Entrevista 73 (ME-258-11H-05).txt"
entrevista <- readLines(archivo, encoding = "UTF-8")
tail(entrevista)

```

---

# Identificación de palabras terminadas en "it[oa]s?"

```{r}

dims <- str_subset(entrevista, "\\w+ \\w+it[oa]s? \\w+") %>%
  str_extract_all("\\w+it[oa]s?\\b")

head(dims, n = 4)

```

---
background-image: url(`r xaringan:::karl`)
background-size: cover
class: center, bottom, inverse

# I was so happy to have discovered RegEx!

---
# Identificación de palabras esdrújulas

Palabras esdrújulas de la forma X.'V.CV.CV(C)

```{r}

esdru <- str_subset(entrevista, "\\w*[áéíóú][bcdfghjklmnprstvxyz]+[aeiou][bcdfghjklmnprstvxyz][aeiou][bcdfghjklmnprstvxyz]?\\b") %>%
  str_extract_all("\\w*[áéíóú][bcdfghjklmnprstvxyz]+[aeiou][bcdfghjklmnprstvxyz][aeiou][bcdfghjklmnprstvxyz]?\\b")

head(esdru, n = 4)

```

---

# De lista a vector

Una vez identificados los elementos relevantes mediante RegEx, se pueden almacenar como vectores y luego como data frames.

```{r}

vec <- vector('character')

for (v in dims) {
  for (s in v) {
    #print(string)
    vec <- c(vec, s)
  }
}

head(vec, n = 5)

```

---

#De vector a data frame

```{r}

word_frame <- as.data.frame(table(vec)) %>%
  arrange(desc(Freq)) %>%
  mutate(vec = fct_reorder(vec, desc(Freq)))

head(word_frame)

```

---

# Descripción cuantitativa

```{r echo=FALSE}

ggplot(word_frame, aes(x = vec, y = Freq)) +
  geom_bar(stat = "identity")

```

---

background-image: url(`r xaringan:::karl`)
background-position: 50% 50%
class: center, bottom, inverse

# So amazing!

---
#Palabras esdrújulas más frecuentes

```{r echo=FALSE}

esdrujulas <- vector("character")

for (v in esdru) {
  for (s in v) {
    esdrujulas <- c(esdrujulas, s)
  }
}

#head(esdrujulas, n = 5)

word_frame2 <- as.data.frame(table(esdrujulas)) %>%
  arrange(desc(Freq)) %>%
  mutate(esdrujulas = fct_reorder(esdrujulas, desc(Freq)))

head(word_frame2)
```

---
#Descripción cuantitativa

```{r echo=FALSE}

ggplot(word_frame2, aes(x = esdrujulas, y = Freq)) +
  geom_bar(stat = "identity")

```


---
class: inverse, middle, center
background-image: url(https://upload.wikimedia.org/wikipedia/commons/3/39/Naruto_Shiki_Fujin.svg)
background-size: contain

# Naruto

---
background-image: url(https://upload.wikimedia.org/wikipedia/commons/b/be/Sharingan_triple.svg)
background-size: 100px
background-position: 90% 8%

# Sharingan

The R package name **xaringan** was derived<sup>1</sup> from **Sharingan**, a dōjutsu in the Japanese anime _Naruto_ with two abilities:

- the "Eye of Insight"

- the "Eye of Hypnotism"

I think a presentation is basically a way to communicate insights to the audience, and a great presentation may even "hypnotize" the audience.<sup>2,3</sup>

.footnote[
[1] In Chinese, the pronounciation of _X_ is _Sh_ /ʃ/ (as in _shrimp_). Now you should have a better idea of how to pronounce my last name _Xie_.

[2] By comparison, bad presentations only put the audience to sleep.

[3] Personally I find that setting background images for slides is a killer feature of remark.js. It is an effective way to bring visual impact into your presentations.
]

---

# Naruto terminology

The **xaringan** package borrowed a few terms from Naruto, such as

- [Sharingan](http://naruto.wikia.com/wiki/Sharingan) (写輪眼; the package name)

- The [moon reader](http://naruto.wikia.com/wiki/Moon_Reader) (月読; an attractive R Markdown output format)

- [Chakra](http://naruto.wikia.com/wiki/Chakra) (查克拉; the path to the remark.js library, which is the power to drive the presentation)

- [Nature transformation](http://naruto.wikia.com/wiki/Nature_Transformation) (性質変化; transform the chakra by setting different options)

- The [infinite moon reader](http://naruto.wikia.com/wiki/Infinite_Tsukuyomi) (無限月読; start a local web server to continuously serve your slides)

- The [summoning technique](http://naruto.wikia.com/wiki/Summoning_Technique) (download remark.js from the web)

You can click the links to know more about them if you want. The jutsu "Moon Reader" may seem a little evil, but that does not mean your slides are evil.

---

class: center

# Hand seals (印)

Press `h` or `?` to see the possible ninjutsu you can use in remark.js.

![](https://upload.wikimedia.org/wikipedia/commons/7/7e/Mudra-Naruto-KageBunshin.svg)

---

class: center, middle

# Thanks!

Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).

The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](http://yihui.name/knitr), and [R Markdown](https://rmarkdown.rstudio.com).
