---
title: "Correlación entre el uso de _ahorita_ y la edad de los hablantes"
#subtitle: "⚔<br/>with xaringan"
#author: "Yihui Xie"
#institute: "RStudio, Inc."
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

background-image: url(https://upload.wikimedia.org/wikipedia/commons/b/be/Sharingan_triple.svg)

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(dplyr)
library(ggplot2)
library(forcats)

```

???

Image credit: [Wikimedia Commons](https://commons.wikimedia.org/wiki/File:Sharingan_triple.svg)

---
class: center, middle

# xaringan

### /ʃaː.'riŋ.ɡan/

---
class: inverse, center, middle

# Get Started

---

# Corpus

Para la investigación se utilizaron las 108 entrevistas del _Corpus Sociolingüístico de la Ciudad de México_.

--
- Se contó cuántas instancias de _ahorita_ y de _ahora_ realizaba cada hablante.

--
- Se registraron los datos sociolingüísticos de los informantes.

```{r}
FrecHablantes <- read.csv("FreqHablantesR.csv")

glimpse(FrecHablantes)

```

---
#Distribución de ahorita por grupos etarios

```{r echo=FALSE}

FrecHablantes$GpoEtario <- factor(FrecHablantes$GpoEtario, levels = c("V", "A", "J"))

ggplot(FrecHablantes, aes(x = GpoEtario, y = Ahorita)) +
  geom_boxplot()

```

---

#Uso de ahorita por grupos etarios + género

```{r echo=FALSE}

FrecHablantes$Genero <- factor(FrecHablantes$Genero, levels = c("M", "H"))

ggplot(FrecHablantes, aes(x = GpoEtario, y = Ahorita, fill = Genero)) +
  geom_boxplot()

```

---

#Uso de ahorita por año de nacimiento

```{r echo=FALSE}

ggplot(FrecHablantes, aes(x = ANacimiento, y = Ahorita)) +
  geom_point()

```

---

#Superposición del modelo lineal

```{r echo=FALSE}

ggplot(FrecHablantes, aes(x = ANacimiento, y = Ahorita)) +
  geom_point() + 
  geom_smooth(method = "lm")

```


---


#Correlación entre las variables

```{r}

FrecHablantes %>%
  summarize(N = n(), r = cor(ANacimiento, Ahorita))

```


---



#Modelo: Frecuencia de uso de ahorita ~ Año de nacimiento de los hablantes

```{r echo=FALSE}

modelo <- lm(Ahorita ~ ANacimiento, data = FrecHablantes)

summary(modelo)

```



---

#Media de uso de ahorita por año de nacimiento

```{r}

fAhorita_year <- FrecHablantes %>%
  group_by(ANacimiento) %>%
  summarise(Media = round(mean(Ahorita), digits = 1))
  
head(fAhorita_year)

```


---

#Uso medio de ahorita por año de nacimiento

```{r echo=FALSE}

ggplot(fAhorita_year, aes(x = ANacimiento, y = Media)) +
  geom_point() + 
  geom_smooth(method = "lm")

```


---
#Correlación entre las variables

```{r}

fAhorita_year %>%
  summarize(N = n(), r = cor(ANacimiento, Media))

```


---


#Modelo: Media de uso de ahorita ~ Año de nacimiento

```{r echo=FALSE}

modelo2 <- lm(Media ~ ANacimiento, data = fAhorita_year)

summary(modelo2)

```



---
class: inverse, middle, center
background-image: url(https://upload.wikimedia.org/wikipedia/commons/3/39/Naruto_Shiki_Fujin.svg)
background-size: contain

# Naruto

---
background-image: url(https://upload.wikimedia.org/wikipedia/commons/b/be/Sharingan_triple.svg)
background-size: 100px
background-position: 90% 8%

# Sharingan

The R package name **xaringan** was derived<sup>1</sup> from **Sharingan**, a dōjutsu in the Japanese anime _Naruto_ with two abilities:

- the "Eye of Insight"

- the "Eye of Hypnotism"

I think a presentation is basically a way to communicate insights to the audience, and a great presentation may even "hypnotize" the audience.<sup>2,3</sup>

.footnote[
[1] In Chinese, the pronounciation of _X_ is _Sh_ /ʃ/ (as in _shrimp_). Now you should have a better idea of how to pronounce my last name _Xie_.

[2] By comparison, bad presentations only put the audience to sleep.

[3] Personally I find that setting background images for slides is a killer feature of remark.js. It is an effective way to bring visual impact into your presentations.
]

---

# Naruto terminology

The **xaringan** package borrowed a few terms from Naruto, such as

- [Sharingan](http://naruto.wikia.com/wiki/Sharingan) (写輪眼; the package name)

- The [moon reader](http://naruto.wikia.com/wiki/Moon_Reader) (月読; an attractive R Markdown output format)

- [Chakra](http://naruto.wikia.com/wiki/Chakra) (查克拉; the path to the remark.js library, which is the power to drive the presentation)

- [Nature transformation](http://naruto.wikia.com/wiki/Nature_Transformation) (性質変化; transform the chakra by setting different options)

- The [infinite moon reader](http://naruto.wikia.com/wiki/Infinite_Tsukuyomi) (無限月読; start a local web server to continuously serve your slides)

- The [summoning technique](http://naruto.wikia.com/wiki/Summoning_Technique) (download remark.js from the web)

You can click the links to know more about them if you want. The jutsu "Moon Reader" may seem a little evil, but that does not mean your slides are evil.

---

class: center

# Hand seals (印)

Press `h` or `?` to see the possible ninjutsu you can use in remark.js.

![](https://upload.wikimedia.org/wikipedia/commons/7/7e/Mudra-Naruto-KageBunshin.svg)

---

class: center, middle

# Thanks!

Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).

The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](http://yihui.name/knitr), and [R Markdown](https://rmarkdown.rstudio.com).
